Modelling the Lymphatic Metastatic Progression Pathways of OPSCC from Multi-Institutional Datasets

### 2023-12-18 src/tex/ms.tex
[showyourwork] overleaf sync

% Abstract
\begin{abstract}
    Objective: The \gls{ctv-n} in \gls{opscc} is currently based mostly on the prevalence of lymph node metastases in different \glspl{lnl} for a given primary tumor location. We present a probabilistic model for ipsilateral lymphatic spread that can quantify the risk for microscopic nodal involvement based on an individual patient's T-category and clinical involvement of \glspl{lnl} at diagnosis.

    Methods: We extend a previously published \gls{hmm}, which models the \glspl{lnl} (I, II, III, IV, V, and VII) as hidden binary \glspl{rv}. Each represents a patient's true state of lymphatic involvement. Clinical involvement at diagnosis represents the observed binary \glspl{rv} linked to the true state via sensitivity and specificity. The primary tumor and the hidden \glspl{rv} are connected in a graph. Each edge represents the conditional probability of metastatic spread per abstract time-step, given involvement at the edge's starting node. To learn these probabilities, we draw \acrlong{mcmc} samples from the likelihood of a dataset (\variable{output/num_patients.txt} \gls{opscc} patients) from three institutions. We compute the model evidence using \acrlong{ti} for different graphs to determine which describes the data best.

    Results: The graph maximizing the model evidence connects the tumor to each \gls{lnl} and the \glspl{lnl} I through V in order. It predicts the risk of occult disease in level IV is below 5\% if level III is clinically negative, and that the risk of occult disease in level V is below 5\% except for advanced T-category (T3 and T4) patients with clinical involvement of levels II, III, and IV.

    Conclusion: The provided statistical model of nodal involvement in \gls{opscc} patients trained on multi-institutional data may guide the design of clinical trials on volume-deescalated treatment of \gls{opscc} and contribute to further personalized guidelines on elective nodal treatment.
\end{abstract}

>>>>

% Abstract
\begin{abstract}
    The \gls{ctv-n} in \gls{opscc} is currently based mostly on the prevalence of lymph node metastases in different \glspl{lnl} for a given primary tumor location. We present a probabilistic model for ipsilateral lymphatic spread that can quantify the microscopic nodal involvement risk based on an individual patient's T-category and clinical involvement of \glspl{lnl} at diagnosis.

    We extend a previously published \gls{hmm}, which models the \glspl{lnl} (I, II, III, IV, V, and VII) as hidden binary \glspl{rv}. Each represents a patient's true state of lymphatic involvement. Clinical involvement at diagnosis represents the observed binary \glspl{rv} linked to the true state via sensitivity and specificity. The primary tumor and the hidden \glspl{rv} are connected in a graph. Each edge represents the conditional probability of metastatic spread per abstract time-step, given disease at the edge's starting node. To learn these probabilities, we draw \acrlong{mcmc} samples from the likelihood of a dataset (\variable{output/num_patients.txt} \gls{opscc} patients) from three institutions. We compute the model evidence using \acrlong{ti} for different graphs to determine which describes the data best.

    The graph maximizing the model evidence connects the tumor to each \gls{lnl} and the \glspl{lnl} I through V in order. It predicts the risk of occult disease in level IV is below 5\% if level III is clinically negative, and that the risk of occult disease in level V is below 5\% except for advanced T-category (T3 and T4) patients with clinical involvement of levels II, III, and IV.

    The provided statistical model of nodal involvement in \gls{opscc} patients trained on multi-institutional data may guide the design of clinical trials on volume-deescalated treatment of \gls{opscc} and contribute to more personal guidelines on elective nodal treatment.
\end{abstract}
